Stochastic modeling of EGS using continuum-fracture approach.

### 2019-07-26 paper.tex

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts,amsmath} % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage[bbgreekl]{mathbbol} % for blackboard \sigma

\newcommand{\CC}{\mathbb C}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\eq}[1]{\begin{equation}#1\end{equation}}
\newcommand{\II}{\mathbb I}
\newcommand{\jmp}[1]{[\![#1]\!]}
\newcommand{\nn}{\vc n}
\newcommand{\uu}{\vc u}
\newcommand{\vc}[1]{\boldsymbol{#1}}
\newcommand{\xx}{\vc x}

\title{Stochastic modeling of EGS using continuum-fracture approach and multilevel Monte Carlo}

@@

\section{Introduction}

\section{Mathematical model}

According to available geological data of the Litoměřice locality, the rock in the depth of several km is considered crystallinic, possibly consisting of granite, gneiss or mica schist \cite{capova}. % zula, rula, svor
The mechanical properties of granitic rocks have been investigated e.g. in \cite{ljunggren}, where the measured Young modulus $E$ varies around 50 GPa and Poisson's ratio $\nu$ is around 0.25.
The reported values of compressibility $\beta$ of several types of granite and orthogneiss are approx. $20\times 10^{-13}$ cm${}^2$.dyne${}^{-1}$ = $2\times10^{-11}$ Pa${}^{-1}$ \cite{zisman}.
As far as porosity $\vartheta$ is concerned, \cite{intera} report values between 0.001 and 0.01 for metamorphic crystalline rocks, we shall consider the value 0.005.
For water we use the density $\varrho_l=10^3$ kg.m${}^{-3}$, compressibility $\beta_l=5\times 10^{-10}$ Pa${}^{-1}$.
Here and in what follows, subscript ``l'' indicates quantities associated to the liquid.
Similarly, subscript ``f'' will denote quantities for the fractures.
Storativity $S$ is calculated from compressibility and porosity as follows (\cite{brace-et-al}, see also \cite{intera}):
\[ S = \varrho_l g(\beta + \vartheta\beta_l). \]
Here $g=9.81$ m.s${}^{-2}$ is the standard gravity.
In our case we obtain $S\approx 2\times 10^{-7} \mbox{ m}^{-1}$.

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
 & rock & fractures & liquid\\\hline
Density $\varrho$ [kg.m${}^{-3}$] & 2700 & 2700 & 1000 \\
Hydraulic conductivity $k$ [m.s${}^{-1}$] & $10^{-9}$ \cite{sperl-trckova} & $10^{-3}$ & ---\\
Porosity $\vartheta$ [--] & $5\times10^{-3}$ & $5\times10^{-2}$ & --- \\
Compressibility $\beta$ [Pa${}^{-1}$] & $2\times10^{-11}$ & $2\times10^{-11}$ & $5\times10^{-10}$ \\
Storativity $S$ [m${}^{-1}$] & $2\times10^{-7}$ & $2\times10^{-7}$ & --- \\
Cross-section $\delta$ [m] & --- & $10^{-3}$ & --- \\
\hline
Young modulus $E$ [Pa] & $50\times10^9$ & 50 & --- \\
Poisson's ratio $\nu$ [--] & 0.25 & 0.25 & --- \\
Biot's coefficient $\alpha$ [--] & 1 & 1 & --- \\
\hline
Heat capacity $c$ [J.kg${}^{-1}$.K${}^{-1}$] & 790 & 790 & 4000 \\
Thermal conductivity $\lambda$ [W.m${}^{-1}$.K${}^{-1}$] & 2.5 & 2.5 & 0.5 \\
\hline
\end{tabular}
\caption{Physical parameters of rock, fractures and water, used in the model.}
\label{tab:params}
\end{table}

@@

In the first model scenario the fluid is injected to both wells in order to open the preexisting fractures.
This leads to increase in the cross-section of fractures, which is then used in the computation of long-term operation.
As the HM model of 3d rock we use the Biot poroelasticity equations:
\eq{\label{eq:biot} \partial_t(Sp + \alpha\div\uu) +\div\vc q = 0,\qquad -\div\bbsigma + \tilde\alpha\nabla p = 0. }
Here $p=h-x_3$ is the pressure head [m], $\uu$ is the displacement [m] and $\tilde\alpha:=\alpha\varrho_l g$.
The flux $\vc q$ [m.s${}^{-1}$] and the stress tensor $\bbsigma$ [Pa] are given by the Darcy and Hooke laws:
\eq{ \vc q = - k\nabla p, \qquad \bbsigma = \CC(\nabla\uu) := \mu(\nabla\uu+\nabla\uu^\top) + \lambda(\II:\nabla\uu)\II, }
where $\mu=E/(2(1+\nu))$ and $\lambda=E\nu/((1+\nu)(1-2\nu))$ are the Lamé parameters [Pa], $\II$ is the identity matrix and ``:'' denotes the scalar product of matrices.

For the 2d fractures we apply a dimension reduction procedure, which was originally developed in \cite{martin-jaffre-roberts} for the Darcy flow.
We introduce the unit normal vector $\vc\nu$ to the fracture plane with arbitrary but fixed orientation.
For a quantity, say $p$, which is defined on both sides of the fracture, we define the jump $\jmp{p}:=p^+-p^-$, where $p^\pm$ is the trace of $p$ on the positive/negative side of the fracture, determined by the orientation of $\vc\nu$.
Then, integrating \eqref{eq:biot} across fracture aperture results in the following system of equations:
\eq{\label{eq:biot_f}\left.\begin{aligned}
\delta\left(\partial_t\left(S_fp_f + \alpha_f\left(\div_\tau\uu_f+\tfrac1\delta\jmp{\uu}\cdot\vc\nu\right)\right) - k_f\Delta_\tau p_f\right) &= \jmp{F}, \\
\delta\left(-\div_\tau\CC_f(\nabla_\tau\uu_f+\tfrac1\delta\vc\nu\otimes\jmp{\uu}) + \tilde\alpha_f\nabla_\tau p_f\right) &= \jmp{\vc Q}.
\end{aligned}\right\} }
Here $p_f$ and $\uu_f$ is the average pressure head and displacement in fractures, the subscript $\tau$ indicates tangential operators in the fracture plane, ``$\otimes$'' denotes the outer product and $\tilde\alpha_f:=\alpha_f\varrho_l g$.
The expressions $\jmp{F}=F^+-F^-$ and $\jmp{\vc Q}=\vc Q^+-\vc Q^-$ are sources and forces that arise from the approximation of derivatives in the direction orthogonal to fracture plane:
\eq{ \left.\begin{aligned}
F^\pm &:= \pm\frac2\delta k_f(p^\pm-p_f),\\
\vc Q^\pm &:= \pm\frac2\delta\CC_f(\nabla_\tau\uu^\pm + \vc\nu\otimes(\uu^\pm-\uu_f))\vc\nu. %\pm\frac2\delta\left(\mu_f(\uu^\pm-\uu_f)+(\mu_f+\lambda_f)((\uu^\pm-\uu_f)\cdot\vc\nu)\vc\nu\right) +\mu_f(\nabla_\tau\uu^\pm)^\top\vc\nu + \lambda_f(\div_\tau\uu^\pm)\vc\nu \pm\tilde\alpha_f p_f\vc\nu.
\end{aligned}\right\} }
Systems \eqref{eq:biot} and \eqref{eq:biot_f} are complemented by the following interface conditions on the fractures:
\eq{ k\nabla p^\pm\cdot\vc\nu = F^\pm,\qquad \left(\CC(\nabla\uu^\pm) - \tilde\alpha p^\pm\II\right)\vc\nu = \vc Q^\pm. }

@@

@@

\begin{thebibliography}{9}

\bibitem{brace-et-al} W. F. Brace,  J. B. Walsh, W. T. Grango.  Permeability of Granite Under High Pressure, J. Geophys. Res. 73(6):2225--2236, 1968.

\bibitem{capova} L. Čápová. Specification of the geothermic model in the environs of several selected boreholes. Diploma thesis, Charles University in Prague, 2013.

\bibitem{intera} INTERA Environmental Consultants, Inc. Porosity, Permeability,  and Their  Relationship  in  
Granite,  Basalt, and Tuff. Accession  DE83-011519, NTIS, Springfield, Virginia, 1983.

\bibitem{martin-jaffre-roberts} V. Martin, J. Jaffré, J. E. Roberts. Modeling fractures and barriers as interfaces for flow in porous media. SIAM Journal on Scientific Computing 26(5):1667--1691, 2005.

\bibitem{sperl-trckova} J. Šperl, J. Trčková. Permeability and porosity of rocks and their relationship based on laboratory testing. Acta Geodyn. Geomater. 5(149):41--47, 2008.

\bibitem{ljunggren} C. Ljunggren, O. Stephansson, O. Alm, H. Hakami, U. Mattila. Mechanical properties of granitic rocks from Gide\aa, Sweden. Technical Report 85-06, SKB, 1985.

\bibitem{zisman} W. A. Zisman. Compressibility and anisotropy of rocks at and near the Earth's surface. Proceedings of the National Academy of Sciences of the United States of America, 19(7):666--679, 1933.

\end{thebibliography}

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage[colorlinks]{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts,amsmath} % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage[bbgreekl]{mathbbol} % for blackboard \sigma
\usepackage{color}
\usepackage[square]{natbib}

\newcommand{\CC}{\mathbb C}
\renewcommand{\div}{\operatorname{div}}
\newcommand{\eq}[1]{\begin{equation}#1\end{equation}}
\newcommand{\II}{\mathbb I}
\newcommand{\jmp}[1]{[\![#1]\!]}
\newcommand{\nn}{\vc n}
\newcommand{\uu}{\vc u}
\newcommand{\vc}[1]{\boldsymbol{#1}}
\newcommand{\xx}{\vc x}
\newcommand{\todo}[1]{{\color{red}#1}}

\title{Stochastic modeling of EGS using continuum-fracture approach.}

@@

\section{Introduction}
\todo{
- problem definition - uncertainty of EGS output due to  fracture network and other properties\\
- goal - capturing these uncertainties by stochastic modelling\\
- basic research (find some related work for dealing with modelling EGS with fracture opening and/or stochastic DFN\\
- paper structure\\
}

\section{Mathematical model}
According to available geological data of the Litoměřice locality, the rock in the depth of several km is considered crystallinic, possibly consisting of granite, gneiss or mica schist \cite{Capova2013}. % zula, rula, svor
We also refer to the EGS study for this site in  contribution \cite{Ralek}.

\todo{short characterization of the locality, rock types,  temperature, reference to Ralek, identification of the model parameters postpone after model introduction.}

\todo{Conceptual model:

\begin{itemize}
    \item modeling just small box around the heat exchanger
    \item fracking considered as just opening existing fractures;simple HM model
    \item deformation considered permanent
    \item longterm heat transfer as separated problem;
          TH model sequentialy coupled
\end{itemize}
}

@@

In the first model scenario the fluid is injected to both wells in order to open the preexisting fractures.
This leads to increase in the cross-section of fractures, which is then used in the computation of long-term operation.
As the HM model of 3d rock we use the Biot poroelasticity equations:
\eq{\label{eq:biot} \partial_t(Sp + \alpha\div\uu) +\div\vc q = 0,\qquad -\div\bbsigma + \tilde\alpha\nabla p = 0. }
Here $p=h-x_3$ is the pressure head [m], $\uu$ is the displacement [m], $\tilde\alpha:=\alpha\varrho_l g$ and $\alpha$ is the Biot's effective stress parameter \cite{Biot1941}.
The flux $\vc q$ [m.s${}^{-1}$] and the stress tensor $\bbsigma$ [Pa] are given by the Darcy and Hooke laws:
\eq{ \vc q = - k\nabla p, \qquad \bbsigma = \CC(\nabla\uu) := \mu(\nabla\uu+\nabla\uu^\top) + \lambda(\II:\nabla\uu)\II, }
where $\mu=E/(2(1+\nu))$ and $\lambda=E\nu/((1+\nu)(1-2\nu))$ are the Lamé parameters [Pa], $\II$ is the identity matrix and ``:'' denotes the scalar product of matrices.

For the 2d fractures we apply a dimension reduction procedure, which was originally developed in \cite{martin_modeling_2005} for the Darcy flow.
We introduce the unit normal vector $\vc\nu$ to the fracture plane with arbitrary but fixed orientation.
For a quantity, say $p$, which is defined on both sides of the fracture, we define the jump $\jmp{p}:=p^+-p^-$, where $p^\pm$ is the trace of $p$ on the positive/negative side of the fracture, determined by the orientation of $\vc\nu$.
Then, integrating \eqref{eq:biot} across fracture aperture results in the following system of equations:
\eq{\label{eq:biot_f}\left.\begin{aligned}
\delta\left(\partial_t\left(S_fp_f + \alpha_f\left(\div_\tau\uu_f+\tfrac1\delta\jmp{\uu}\cdot\vc\nu\right)\right) - k_f\Delta_\tau p_f\right) &= \jmp{F}, \\
\delta\left(-\div_\tau\CC_f(\nabla_\tau\uu_f+\tfrac1\delta\vc\nu\otimes\jmp{\uu}) + \tilde\alpha_f\nabla_\tau p_f\right) &= \jmp{\vc Q}.
\end{aligned}\right\} }
Here $p_f$ and $\uu_f$ is the average pressure head and displacement in fractures, the subscript $\tau$ indicates tangential operators in the fracture plane, ``$\otimes$'' denotes the outer product and $\tilde\alpha_f:=\alpha_f\varrho_l g$.
The expressions $\jmp{F}=F^+-F^-$ and $\jmp{\vc Q}=\vc Q^+-\vc Q^-$ are sources and forces that arise from the approximation of derivatives in the direction orthogonal to fracture plane:
\eq{ \left.\begin{aligned}
F^\pm &:= \pm\frac2\delta k_f(p^\pm-p_f),\\
\vc Q^\pm &:= \pm\frac2\delta\CC_f(\nabla_\tau\uu^\pm + \vc\nu\otimes(\uu^\pm-\uu_f))\vc\nu. %\pm\frac2\delta\left(\mu_f(\uu^\pm-\uu_f)+(\mu_f+\lambda_f)((\uu^\pm-\uu_f)\cdot\vc\nu)\vc\nu\right) +\mu_f(\nabla_\tau\uu^\pm)^\top\vc\nu + \lambda_f(\div_\tau\uu^\pm)\vc\nu \pm\tilde\alpha_f p_f\vc\nu.
\end{aligned}\right\} }
Systems \eqref{eq:biot} and \eqref{eq:biot_f} are complemented by the following interface conditions on the fractures:
\eq{ k\nabla p^\pm\cdot\vc\nu = F^\pm,\qquad \left(\CC(\nabla\uu^\pm) - \tilde\alpha p^\pm\II\right)\vc\nu = \vc Q^\pm. }

@@

\subsection{Conductivity dependent on displacement}
We consider that after releasing the fracking pressure the displacement of larger fractures persists. So the change in new porosity is
\[
\phi = \phi_0 + \alpha\div \vc u
\]
where $\phi_0$ is the porosity before deformation, c.f. time term in \eqref{eq:biot}. 
Using the Carman-Kozeny relation \cite{Kaviany1999}, 
% \[
%     k = \tilde{k}\frac{\phi^3}{(1-\phi)^2}
% \]
we can derive formula for the updated conductivity:
\[
    K = K_0\left(\frac{\phi}{\phi_0}\right)^3 \left(\frac{1-\phi_0}{1-\phi}\right)^2.
\]

The mechanical properties of granitic rocks have been investigated e.g. in \cite{Ljunggren1985}, where the measured Young modulus $E$ varies around 50 GPa and Poisson's ratio $\nu$ is around 0.25.
The reported values of the drained compressibility $\beta$ of several types of granite and orthogneiss are approx. $20\times 10^{-13}$ cm${}^2$.dyne${}^{-1}$ = $2\times10^{-11}$ Pa${}^{-1}$ \cite{Zisman1933}.
As far as porosity $\vartheta$ is concerned, \cite{intera} report values between 0.001 and 0.01 for metamorphic crystalline rocks, we shall consider the value 0.005.
For water we use the density $\varrho_l=10^3$ kg.m${}^{-3}$, compressibility $\beta_l=5\times 10^{-10}$ Pa${}^{-1}$.
Here and in what follows, subscript ``l'' indicates quantities associated to the liquid.
Similarly, subscript ``f'' will denote quantities for the fractures.
Storativity $S$ is calculated from compressibility and porosity as follows (\cite{Brace1968}, see also \cite{intera}):
\[ S = \varrho_l g(\beta + \vartheta\beta_l). \]
Here $g=9.81$ m.s${}^{-2}$ is the standard gravity.
In our case we obtain $S\approx 2\times 10^{-7} \mbox{ m}^{-1}$.

\begin{table}
\centering
\begin{tabular}{|l|c|c|c|}
\hline
 & rock & fractures & liquid\\\hline
Density $\varrho$ [kg.m${}^{-3}$] & 2700 & 2700 & 1000 \\
Hydraulic conductivity $k$ [m.s${}^{-1}$] & $10^{-9}$ \cite{Sperl2008} & $10^{-3}$ & ---\\
Porosity $\vartheta$ [--] & $5\times10^{-3}$ & $5\times10^{-2}$ & --- \\
Compressibility $\beta$ [Pa${}^{-1}$] & $2\times10^{-11}$ & $2\times10^{-11}$ & $5\times10^{-10}$ \\
Storativity $S$ [m${}^{-1}$] & $2\times10^{-7}$ & $2\times10^{-7}$ & --- \\
Cross-section $\delta$ [m] & --- & $10^{-3}$ & --- \\
\hline
Young modulus $E$ [Pa] & $50\times10^9$ & 50 & --- \\
Poisson's ratio $\nu$ [--] & 0.25 & 0.25 & --- \\
Biot's coefficient $\alpha$ [--] & 1 & 1 & --- \\
\hline
Heat capacity $c$ [J.kg${}^{-1}$.K${}^{-1}$] & 790 & 790 & 4000 \\
Thermal conductivity $\lambda$ [W.m${}^{-1}$.K${}^{-1}$] & 2.5 & 2.5 & 0.5 \\
\hline
\end{tabular}
\caption{Physical parameters of rock, fractures and water, used in the model.}
\label{tab:params}
\end{table}

\section{Fractured media}
\subsection{Stochastic description of the fractures}
\subsection{Effective properties of the fractured porous media}

@@

\bibliographystyle{dinat}
\bibliography{references.bib}

% \begin{thebibliography}{9}

% \bibitem{brace-et-al} W. F. Brace,  J. B. Walsh, W. T. Grango.  Permeability of Granite Under High Pressure, J. Geophys. Res. 73(6):2225--2236, 1968.

% \bibitem{capova} L. Čápová. Specification of the geothermic model in the environs of several selected boreholes. Diploma thesis, Charles University in Prague, 2013.

% \bibitem{intera} INTERA Environmental Consultants, Inc. Porosity, Permeability,  and Their  Relationship  in  
% Granite,  Basalt, and Tuff. Accession  DE83-011519, NTIS, Springfield, Virginia, 1983.

% \bibitem{martin-jaffre-roberts} V. Martin, J. Jaffré, J. E. Roberts. Modeling fractures and barriers as interfaces for flow in porous media. SIAM Journal on Scientific Computing 26(5):1667--1691, 2005.

% \bibitem{sperl-trckova} J. Šperl, J. Trčková. Permeability and porosity of rocks and their relationship based on laboratory testing. Acta Geodyn. Geomater. 5(149):41--47, 2008.

% \bibitem{ljunggren} C. Ljunggren, O. Stephansson, O. Alm, H. Hakami, U. Mattila. Mechanical properties of granitic rocks from Gide\aa, Sweden. Technical Report 85-06, SKB, 1985.

% \bibitem{zisman} W. A. Zisman. Compressibility and anisotropy of rocks at and near the Earth's surface. Proceedings of the National Academy of Sciences of the United States of America, 19(7):666--679, 1933.

% \end{thebibliography}
